Parameter-less Population Pyramid

### 2014-01-05 Parameterless Population Pyramid/main.tex
Minor word updates

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal often requires an optimizer to make
assumptions about the nature of the problem is is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.

@@

For our purposes, we use a First Improvement Hill Climber
(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
  various names in the literature, but use it here with a descriptive
  name for clarity.}, a method of exploiting local information through
a series of modify-evaluate-accept iterations starting from a randomly
initialized solution and ending with a optimum, either local or global.
FIHC uses the following algorithm.  Consider all gene loci
$x_i$ in a random order.  For a given $x_i$ try all possible
alternative values for the gene $x_i$, evaluating solution quality
with change.  If any alternative value is a strict fitness improvement
over the current value for $x_i$, replace $x_i$ with the improved
value.  Iteration of the process continues until no such change results in
a strict fitness improvement.

@@

Starting from the leaves, the binary tree can be constructed using an
agglomerative method.  Each
iteration determines the pairwise linkage between each cluster of loci
using Equation~\ref{eq-distance}, and merge the two closest into a new
cluster.  The process ends when there is only a single cluster
containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

@@

As P3 has exploited all known information each time the top of the pyramid is reached,
there no wasted evaluations on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence. It is this essential balance of
evaluation to diverstiy that P3 focuses on.

@@

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied.  Each causes their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a crossover.  As such we currently are unable to state conclusively
that updating $P_i$ will require $O(N)$ evaluations.  That said,
empirically it to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

@@

\section{Promising Theory}
Due to P3's persistent integration of random solutions, it always has
a non zero probability of reaching the global optimum.  Because of
this (and unlike LTGA), it is possible to derive expected
running times for P3 on a number of problems.  While we do not present
a rigorous solution to any particular problem here, though we do wish to
illuminate potential avenues for future proofs.

>>>>

\section{Introduction}
Providing a good solution in reasonable time to a broad class of real
world black box problems is a fundamental goal of evolutionary
computation.  Achieving this goal often requires an optimizer to make
assumptions about the nature of the problem it is solving as no
optimizer can solve all problems equally well~\cite{Wolpert:1997:nfl}.

@@

For our purposes, we use a First Improvement Hill Climber
(FIHC)\footnote{We use the name FHIC for an algorithm that likely has
  various names in the literature, but use it here with a descriptive
  name for clarity.}, a method of exploiting local information through
a series of modify-evaluate-accept iterations starting from a randomly
initialized solution and ending with a optimum, either local or global.
FIHC uses the following algorithm.  Consider all gene loci
$x_i$ in a random order.  For a given $x_i$ try all possible
alternative values for the gene $x_i$, evaluating solution quality
with each change.  If any alternative value is a strict fitness improvement
over the current value for $x_i$, replace $x_i$ with the improved
value.  Iteration of the process continues until no such change results in
a strict fitness improvement.

@@

Starting from the leaves, the binary tree can be constructed using an
agglomerative method.  Each
iteration determines the pairwise linkage between each cluster of loci
using Equation~\ref{eq-distance}, and merges the two closest into a new
cluster.  The process ends when there is only a single cluster
containing all loci.  As our clustering algorithm is both convex and
satisfies the swapping lemma~\cite{gronau:2007:upgma}, this algorithm
requires $O(N^2)$ total time to correctly construct the linkage tree
from the frequency table.

@@

As P3 has exploited all known information each time the top of the pyramid is reached,
no evaluations are wasted  on unused diversity.  Similarly, because P3
does not throw away previously optimized solutions while adding diversity, it avoids
waste caused by premature convergence. It is this essential balance of
evaluation to diverstiy that P3 focuses on.

@@

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied.  Each causes their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost per update to $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a crossover.  As such we currently are unable to state conclusively
that updating $P_i$ will require $O(N)$ evaluations.  That said,
empirically it appears to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

@@

\section{Promising Theory}
Due to P3's persistent integration of random solutions, it always has
a non zero probability of reaching the global optimum.  Because of
this (and unlike LTGA), it is possible to derive expected
running times for P3 on a number of problems.  While we do not present
a rigorous solution to any particular problem here, we wish to
illuminate potential avenues for future proofs.
